Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-08 sources/bmc_template_submission/bmc_article.tex
first submission final

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to capture
the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of biological data across databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019} among others.
By using the existing OpenTree taxonomy programmatic tools to map tip names,
Physcraper has a framework for connecting updated phylogenies data from any
biological database.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for many of these taxa are not available \cite{mctavish_how_2017}.
OpenTree's comprehensive tree of life comprises 2.3 million tips, of which around 90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed in the tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life that, but this data either has not been analysed or the analyses
have not been made publicly available \cite{mctavish_how_2017}.

Assembling a DNA alignment from such a massive database can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been developed and used to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

Physcraper improves on previous work in automating phylogenetics by leveraging
the power of existing homology hypotheses that taxon specialists have assessed
and deemed appropriate for a specific phylogenetic scope.
There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as ``jump-start'' alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
link phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, and data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.

@@

The general Physcraper framework (Figure 1) consists of 4 steps: 1)identifying and
processing a tree and its underlying alignment; 2) performing a BLAST search of
DNA sequences from original alignment on GenBank, and filtering of new sequences;
3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic
analysis and comparing the updated tree to existing phylogenies.

@@

Physcraper has the added advantage of linking taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data that
has never been
incorporated into any phylogenetic estimate provides the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases,
with the main goal of building on published alignments and extending existing
phylogenetic inferences with more data and taxa.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of biological data across databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019} among others.
By using the existing OpenTree taxonomy programmatic tools to map tip names,
Physcraper has a framework for connecting updated phylogenies with data from any
biological database.

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for many of these taxa are not available \cite{mctavish_how_2017}.
OpenTree's comprehensive tree of life comprises 2.3 million tips, of which around
90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed
in the tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life, but this data either has not been analysed or the analyses
have not been made publicly available \cite{mctavish_how_2017}.

Assembling a DNA alignment from such a massive database as GenBank can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been developed and used to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper, pearse2013phylogenerator}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

Physcraper improves on previous work in automating phylogenetic reconstruction by leveraging
the power of existing homology hypotheses that taxon specialists have assessed
and deemed appropriate for a specific phylogenetic scope.
There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as ``jump-start'' alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

Physcraper is implemented as a Python pipeline that uses OpenTree's programmatic access
protocols (API's) to automatically
link any phylogeny mapped to OpenTree's standardized taxonomy \cite{mctavish2015phylesystem},
to alignments
from TreeBASE \cite{sanderson1994treebase}, and data from GenBank \cite{benson2000genbank}.
Its utility and functionalities are presented with a case-study analysis of a
group of flowering plants, the hollies.

@@

The general Physcraper framework (Figure 1) consists of 4 steps: 1) identifying and
processing a tree and its underlying alignment; 2) performing a BLAST search of
DNA sequences from original alignment on GenBank, and filtering of new sequences;
3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic
analysis and comparing the updated tree to existing phylogenies.

@@

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
